Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-27 analysis/paper/paper.Rmd
finished editing conclusion

Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. We find that foreign influence may have been stronger in the Chinese period than the European period, contrary to previous work <!-- add citation to previous work -->. The hypothesis of an emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. <!-- add more detail about why this hypothesis was not fully supported --> Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base. Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf. @Acabado2017]. In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct. Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information --> . This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.

In addition to the emergence of specialized groups indicative of changes in social organization, a further explanation could be that producing homogeneous pottery is an act of resistance to the Europeans intrusion and the Chinese immigrants by emphasizing the practice of traditional pottery production. In a culture contact situation, social identity could be expressed through materials practices to show cultural homogeneity and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, not just a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. The case study of Shamaoshan cemetery (3BC- 4AD) in Southwest China showing the process of incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019]. At Kiwulan, the standardization of potter shape is significant in the 19th century when a a large number of Chinese immigrants migrated to Yilan. Shape can be viewed as a symbol as an expression the indigenous identity or social boundary because shape is the most obvious and visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics over time and reached high quantity and diversity in the 19th century, the production of the local pottery continued and became more standardized. This might imply not only the utilitarian function, but an deliberate action to emphasize the pottery tradition as their cultural custom.

@@

This case study demonstrates the application of Elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization and the underlying mechanism indicative of foreign cultural influences. Compared with the traditional linear measurement approach, geometric morphometrics methods especially EFA-based outline approach is more effective to detect shape variations for artifact shape without obvious landmarks. Here, EFA approach combined with a significance test for the equality of CVs for shape variables provides a robust way to identify and statistically assess the difference in shape standardization. Lower variation in ceramic shape was identified after the European presence and even much lower variation during the Chinese presence. The direct relationship between foreign influences and highly standardized ceramic shape was tested here on Kiwulan pottery. This finding helps to understand the cause and factor that would lead to the standardization of pottery production in a cross-cultural interaction context. This gave an insight into the discussion of ceramic specialization which is usually associated with the changes in social organization. The more homogeneous shape during the contact period without any changes in clay paste or producing technique suggests that shape standardization is an outcome resulted from an intentional act. The results here suggest an expression of social identity or showing cultural boundaries in indigenous societies through visible variables when encountering foreign intrusion. The analysis allows comparisons between sites with a similar historical trajectory in a pericolonial context. Our case study includes an openly available pipeline suitable for other assemblages, ensuring reproducibility and expanding the use of shape-based quantitative methods to questions about craft specialization and standardization in prehistoric technologies.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.

>>>>

Our results support the hypothesis that foreign influences at Kiwulan influenced emergence of social inequality in the local Indigenous society. We find that foreign influence may have been stronger in the Chinese period than the European period, contrary to previous work <!-- add citation to previous work -->. The hypothesis of an emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. <!-- add more detail about why this hypothesis was not fully supported --> Compared to other regions in Taiwan, European colonial influence was weak in Yilan due to the isolated of the surrounding mountains and the economic focus of the Spanish and  Dutch who prefered northern Taiwan as their trading base. Indigenous communities Yilan experienced indirect influence from European trade networks and their colonial activities in a pericolonial context [cf. @Acabado2017]. In contrast to the Indigenous-European interactions at Kiwulan, the interaction between indigenous people and  Chinese in the 19th century appears to have been more intense and direct. Historical records indicate that  Chinese immigrants settled in Yilan and lived closely with Kiwulan Indigenous societies <!-- add citation that is the source of this information -->. This direct influence is reflected by the archaeological evidence of large amounts of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Similarly, burials at Kiwulan in this later phase show the adoption of coffins in mortuary practices, which @Chen2007 interprets as the adoption of a symbol of ethnic Chinese.

We recognise that the shape variation reported here is subtle, and this invites consideration of the possibility that the absence of major changes in vessel shape at Kiwulan may have been an act of resistance to foreign. Continuities in vessel shape over time draws our attention to the endurance of traditional pottery production practices amid intrusions from Europeans and Chinese. In a culture contact situation, social identity may be expressed through material practices to show cultural homogeneity and distinction from other groups [@Voss2005]. It is also important to recognize that social identity might be more complicated in a colonial context, and may be more than a colonized–colonizer or local/foreign dichotomy [@Voss2005; @Voss2008]. Shamaoshan cemetery (3BC- 4AD) in Southwest China shows that the process of the incorporation of Southwest China into the Han Empire involved a century of conflicts, resistance, and acceptance among social groups with different identities, especially in the historical context of Han immigrants [@Wu2019]. A similar dynamic may have occured at Kiwulan, with vessel shape indicating both acceptance of forign influence through increased shape standardisation, and resistence through the overall continuity in vessel shape. Vessle shape can be viewed as a symbol as an expression the Indigenous identity and social boundaries because shape is a highly visible trait compared with other features of pottery [cf. @Roux2015]. Although there is an increase in number of imported ceramics over time at Kiwulan, the production of the local pottery continued and became more standardized. This might imply not only a utilitarian function of the local vessels, but a deliberate action to emphasize the local pottery tradition as their cultural custom as foreign contact intensified.

@@

This study demonstrates an application of elliptic Fourier analysis on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust way to identify and statistically assess differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on pottery from Kiwulan, and large Iron Age Indienous settlement in northeast Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to understand the factors that may lead to standardization of pottery production in a cross-cultural interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggests that shape standardization was intentional. The results here further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign intrusion in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington. <!-- add thanks to QRC for funding for residue analysis, give lots of detail -->
